Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-07-08 manuscript/manuscript.tex
Remove entropy eq and add more description to SAS

\subsection{Solvent Accessible Surface (SAS)}
The solvent accessible surface (SAS) represents the interface between regions that are occupied by the solute molecule, and those which can be accessed by the solvent.
GIST provides a very natural way of creating a SAS using an isosurface of the oxygen density \code{gO}.
It can be visualized in PyMOL. 
Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
The expected output is shown in Figure~\ref{fig-streptavidin_sasa}.

@@

\begin{equation}
g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
\end{equation}
A first nearest neighbor (NN) approach is used to evaluate the density.
The densities are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
The average entropy contributions per solvent molecule in voxel $k$ are calculated as

@@

\subsection{Solvation Energy}
The solvation energy is calculated from the water-water and water-solute interactions from the force field.
\begin{equation}
	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
\end{equation}
The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute molecule.
In practice, it is evaluated as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.

@@

\begin{equation}
	\Delta E^\text{R}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
\end{equation}
When the region $R$ comprises more than one solvent molecule, interactions within this region are double-counted while interactions to the outside are not.
This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
\begin{equation}
	\Delta E^\text{R}_{ww} = 2\left(\sum_{k}^\text{voxels in R} E_{ww,k}^\text{corr}-\sum_{k}^\text{voxels in R}\sum_{l}^\text{l>k} E_{ww, k\;l}\right)
\end{equation}
While this is supported by the standard GIST implementation (using the \inlinecode{doeij} flag), it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
This depends on the local environment and on \emph{what} the solvent would be replaced by.
Therefore, the effect of reorganization must be judged for each case individually.
One way to treat reorganization is an end-state approach as shown in the tutorial: all states before and after solvent reorganization are considered. In the tutorial, this corresponds to the solvation of biotin and streptavidin seperated (initial states) and in complex (end-state). Note that all solvent contributions need to be considered to get thermodynamically accurate results in such a case. In practice this therefore requires the integration over the whole grid or at least all voxels with solvent properties different from bulk (i.e. setting a large cutoff).
\subsubsection{PME energy}

In the original version of GIST, the energies are calculated based on the the minimum image convention.
In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.
The Lennard-Jones part $E_\text{lj}$ is computed separately in direct space.

>>>>

\subsection{Solvent Accessible Surface (SAS)}
The solvent accessible surface (SAS) represents the interface between regions that are occupied by the solute molecule, and those which can be accessed by the solvent.
GIST provides a very natural way of creating a SAS using an isosurface of the oxygen density \code{gO}.
At a very low isovalue, an isosurface of the water's oxygen or hydrogen shows areas the water can reach, essentially creating a SAS. 
Here, the SAS was generated from a simulation of water around streptavidin without biotin, leaving the binding site solvent-accessible.
The expected output is shown in Figure~\ref{fig-streptavidin_sasa}.

@@

%\begin{equation}
%g_\textit{vox} \left( \textbf{r}, \omega \right) \approx g_\textit{vox}(\textbf{r}) g_\textit{vox}(\omega)
%\end{equation}
A first nearest neighbor (NN) approach is used to evaluate the entropy based on the orientational and translational distributions.
The distributions are computed relative to a homogeneous distribution of solvent molecules, given a bulk density $\rho^0$.
The average entropy contributions per solvent molecule in voxel $k$ are calculated as

@@

\subsection{Solvation Energy}
The solvation energy is calculated from the water-water and water-solute interactions from the force field.
\begin{equation}
	\Delta E_\textit{solv} = \Delta E_\textit{sw} + \Delta E_\textit{ww}
\end{equation}
The solute-solvent energy $E_{sw}$ can be expressed in terms of the solvent density and the potential $U_{sw}(\textbf{r},\omega)$ induced by the solute molecule.
In practice, $E_{sw}$ is computed as the expectation value $\langle\cdot\rangle$ of the pairwise force field energy $U_{ij}$ between all solvent molecules in voxel $k$ and all solute atoms.

@@

\begin{equation}
	\Delta E^\text{R}_{ww} = \sum_{k}^{\text{voxels in }R} E_{ww,k}^\text{corr}
%	\Delta E^\textit{R,norm}_\textit{ww} = E^\textit{R,norm}_\textit{ww} - 2E^\textit{bulk}_\textit{ww}
\end{equation}
When the region $R$ comprises more than one solvent molecule, interactions within this region are double-counted while interactions to the outside are not.
This could be solved using the solvent-solvent energy between each pair of voxels $E_{ww,k\;l}$:
\begin{equation}
	\Delta E^\text{R}_{ww} = 2\left(\sum_{k}^\text{voxels in R} E_{ww,k}^\text{corr}-\sum_{k}^\text{voxels in R}\sum_{l}^\text{l>k} E_{ww, k\;l}\right)
\end{equation}
While this is supported by the standard GIST implementation (using the \inlinecode{doeij} flag), it is rarely done due to the large size of the $E_{ww,k\;l}$ matrix.
Integrating GIST values over the whole grid corresponds to a process where all the solvent is removed into bulk.
However, when integrating over a small region, this process is less well-defined, since it is unclear to what extent the remaining solvent would reorganize.
This depends on the local environment and on \emph{what} the solvent would be replaced by.
Therefore, the effect of reorganization must be judged for each case individually.
One way to treat reorganization is an end-state approach as shown in the tutorial: all states before and after solvent reorganization are considered. In the tutorial, this corresponds to the solvation of biotin and streptavidin seperated (initial states) and in complex (end-state). Note that all solvent contributions need to be considered to get thermodynamically accurate results in such a case. In practice this therefore requires the integration over the whole grid or at least all voxels with solvent properties different from bulk (i.e. setting a large cutoff).
\subsubsection{PME energy}
In the original version of GIST, the energies are calculated based on the the minimum image convention.
In PME-GIST, the electrostatic energy $E_\text{elec}$ is calculated using the particle mesh Ewald (PME) method, which yields energies that are highly consistent with the Amber MD engine \cite{Chen2021}.
The Lennard-Jones part $E_\text{lj}$ is computed separately in direct space.
